Combining Static and Dynamic Analyses to Secure Programs against Integer Overflows

### 2012-08-31 doc/Manuscript/CGO/paper.tex
Revised Raphael's section, adding rules for static analysis.

\title{An Algorithm to Prune Integer Overflow Tests}
%\subtitle{A case study of the Dinamica EGO toolkit}

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at this level because we want to couple our library with static
analyses, such as the algorithm that we describe in Section~\ref{sec:range}.
% TODO: add text about the interaction between static an dynamic analysis at
% the end of this section.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the conditions that we use to
detect overflows.

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                  & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \vee$ \\
                  & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation.}
\end{figure}

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary to us, because we check integer overflows in
multiplication via the inverse operation of integer division.
Thus, the test prevents a division by zero from happening.
The semantics of a \textsc{Trunc} instruction $x = \ \downarrow_n \ o_1$ is
to assign to $x$ the $n$ least significant bits of $o_1$.
The dynamic checking in this case consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation with loss of precision, even
if it is a benign one.
However, we give users the possibility of disabling checks on truncation.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target programs.
The instrumentation of all the instructions works
in the same way: first, the pass inserts instructions to gather the
information needed to do the assertion.
Then, the pass inserts compare instructions to generate a boolean value
that indicates if there is an overflow or not. 
After this, the pass inserts a conditional branch: 
the false clause of the branch, where the control goes when there is no 
overflow, leads to the first instruction after the instruction
which has received the instrumentation.
the true clause of the branch leads to a block where we do the
overflow handling.

%What we can do in the overflow handling
Our library has three options to handle the overflows. 
First option: Do nothing and ignore the overflows. This option
was used to do our experiments and verify the slowdown produced 
by the new instructions.
Second option: Give a message in stderr whenever an overflow occurs.
This is the standard option. The instrumented program remains with 
the same behavior, but whenever an overflow occurs, the program
prints \texttt{Overflow occurred in FileName.cpp, line X.} in the
standard error stream.
Third option: Abort the execution when an overflow occurs.
This is the more aggressive option. With this option enabled,
whenever an overflow occurs, the program is terminated, not allowing
the execution of the program with undefined behavior due to 
integer overflows.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/TimeCorr}
\end{center}
\caption{\label{fig:TimeCorr}
Correlation between program size (number of var nodes in constraint
graphs after inlining) and analysis runtime (ms).
Coefficient of determination = 0.967.
}
\end{figure}

>>>>

\title{Combining Static and Dynamic Analyses to Secure Programs against
Integer Overflows}
%\subtitle{A case study of the Dinamica EGO toolkit}

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at the intermediate representation level to be able to couple our
library with static analyses, such as the algorithm that we described in
Section~\ref{sec:range}.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
detect overflows.
This figure also shows how the static analysis helps us to avoid inserting
some of those tests.
We only generate the dynamic tests on the right if the conditions in the middle
are false.

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{lll}
Instruction & Static Check & Dynamic Check \\ \\
$x = o_1 \ +_s \ o_2$ & $\lb{I[o_1]} + \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                  & & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $\ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $\lb{I[o_1]} - \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} - \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $\lb{I[o_1]} > \ub{I[o_2]}$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $\mbox{MIN\_INT} < \{ \lb{I[o_1]} \times \lb{I[o_2]}, \lb{I[o_1]} \times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\
                       & $\mbox{MIN\_INT} < \{ \ub{I[o_1]}\times \lb{I[o_2]}, \ub{I[o_1]}\times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & \\ \\
$x = o_1 \ \qside \ n$ & $2^n \times \ub{I[o_1]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$  & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation. If the static check is true, we do not insert the dynamic
check. The values \{MAX/MIN\}\_INT depend on the type of the instruction.}
\end{figure*}

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few interesting cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary to us, because we check integer overflows in
multiplication via the inverse operation of integer division.
Thus, the test prevents a division by zero from happening.
The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
the $n$ least significant bits of $o_1$.
The dynamic checking in this case consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation that might result in
data loss, even if this loss is benign.
To make the dynamic checks more liberal, we give users the possibility of
disabling tests over truncations.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target program.
Although the dynamic check depends on the instruction that is instrumented,
the general modus operandi is the same.
Our dynamic tests check for overflows after they happen.
The code that we insert to detect the overflow diverts the program flow in case
such an event takes place.

% TODO: add a snapshot of a function with, and without the integer overflow
% instrumentation. Let this function be very simple, e.g.,
% int foo(int x, int y) {return x + y; }

Which actions are performed once the overflow is detected depends on the
user.
Our library gives the user three options to handle them. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/TimeCorr}
\end{center}
\caption{\label{fig:TimeCorr}
Correlation between program size (number of var nodes in constraint
graphs after inlining) and analysis runtime (ms).
Each point represents a benchmark.
Coefficient of determination = 0.967.
}
\end{figure}
